Direction estimation by minimum squared arc length

### 2012-01-13 papersm.tex
minor fixes.
added newest version of paper

\author{Robby~G.~McKilliam, Barry G. Quinn and I. Vaughan L. Clarkson%
  \thanks{A preliminary version of some of this material is contained in Chapters 5 and 6 of Robby McKilliam's PhD thesis \cite{McKilliam2010thesis}. Robby~McKilliam is with the Institute for Telecommunications Research, The University of South Australia, SA, 5095. Vaughan~Clarkson is with the School of
     Information Technology \& Electrical Engineering, The University
     of Queensland, QLD., 4072, Australia. Barry~Quinn is with the Department of Statistics, Macquarie
     University, Sydney, NSW, 2109, Australia}
     }
% The paper headers 
\markboth{Direction estimation by minimum squared arc length}{DRAFT \today}

@@

\begin{abstract}
Circular statistics has found substantial application in science and engineering. One of the fundamental problems in circular statistics is that of estimating the \emph{mean direction} of a circular random variable from a number of observations. The standard approach in the literature is called the \emph{sample circular mean} and its asymptotic properties are well known. It can also be computed efficiently in a number of arithmetic operations that is linear in the number of observations. In this paper we consider an alternative estimator called the \emph{angular least squares estimator} that is based on minimising squared arc length. We show how this estimator can be computed efficiently in a linear number of operations using an algorithm from algebraic number theory and we derive its asymptotic properties. We show both theoretically and by simulation that in some scenarios the angular least squares estimator is statistically more accurate than the sample circular mean. As such the results in this paper potentially have implications for the wide variety of fields in science, engineering and statistics that currently use the sample circular mean.
\end{abstract}

\begin{IEEEkeywords}
Circular statistics, mean direction estimation, lattice theory, nearest lattice point problem.
\end{IEEEkeywords}

@@

The field of circular statistics aims to describe the nature of data that is measured in angles or 2-dimensional unit vectors or complex numbers on the unit circle \cite{Mardia_directional_statistics,Jupp_mardia_unified_directional_statistics_1989,mardia_stat_dir_data_jorver_1975,mardia_stat_dir_data_book_1975,Fisher1993,Jammalamadaka_dir_stat_book}.  Such data occur frequently in science, particularly in astronomy, biology \cite{Morellato_circ_stat_phenology_2010,batschelet1981circular,Cochranmigbridcalcircstat2004,Bolesspinylobnavcircstat2003}, medicine \cite{Mann_circstat_med_cement_2003,Lediseasecircstat2003}, geology, geography and meteorology~\cite{CAPACCIONI_imanal_circ_stat_geol_1997,Fullermagfieldrevcircstat1996,Bowersdirstatwindwaves2000,Fisher_bootstrap_geo_1_1990,Marida_geosciences_1981}, and also in engineering, particularly in communications and radar \cite{Fogel1989_bit_synch_zero_crossings,Fogel1988,Elton_circstat_radar_pulse_1994,Elton96robustparameter1996,Clarkson2007,McKilliamLinearTimeBlockPSK2009,McKilliam2007,Clarkson1996}.  A meteorological example is the direction of the wind, and a biological example is the direction of flight taken by a bird. %The field of circular statistics is in some sense an ancient one, probably begun when mankind first started recording the motion of the sun, the moon and the stars.  A thorough historical account of the subject is given by Fisher \cite[Ch.~1]{Fisher1993}.

A fundamental and useful problem in circular statistics is that of estimating the \emph{mean direction} of a circular random variable from a number, say $N$, of observations. For example, if you listen to the weather report you may be told (an estimate of) the direction of the wind.  Obtaining an accurate estimate requires a method for accurately estimating the mean wind direction from a number of observations of the wind direction. The problem of estimating mean direction also appears in signal processing problems such as phase and frequency estimation \cite{Quinn2009_dasp_phase_only_information_loss,Lovell1991,McKilliamFrequencyEstimationByPhaseUnwrapping2009,Quinn2001,Clarkson1999,Tretter1985,McKilliam2010thesis}, polynomial phase estimation~\cite{McKilliam2009asilomar_polyest_lattice,McKilliam2010thesis,Kitchen_polyphase_unwrapping_1994,Morelande_bayes_unwrapping_2009_tsp}, noncoherent detection \cite{McKilliamLinearTimeBlockPSK2009,McKilliam2010thesis,Sweldens2001,Mackenthun1994} and delay and period estimation from sparse observations \cite{Elton96robustparameter1996,Clarkson2007,McKilliam2007,McKilliam2010thesis,Fogel1988,Fogel1989_bit_synch_zero_crossings,Sidiropoulos2005}.

Unlike the case of random variables on the real line where the mean is uniquely defined, the definition of \emph{mean direction} of a circular random variable is somewhat open to interpretation. The most common definition in the literature, called the \term{circular mean}, is derived by considering circular random variables as unit vectors in 2-dimensions or equivalently complex numbers on the unit circle. Given $N$ observations the appropriate estimator for the circular mean is the \emph{sample circular mean} which can be computed efficiently by averaging $N$ complex numbers. This requires only $O(N)$ arithmetic operations.  The asymptotic properties of the sample circular mean under various assumptions are also well studied \cite{Mardia_directional_statistics,Quinn2009_dasp_phase_only_information_loss,Jammalamadaka_dir_stat_book,McKilliam2010thesis}.

In this paper we consider an alternative definition of mean direction called the \term{unwrapped mean}\index{unwrapped mean} which is based on minimising the expected squared arc length. This definition is similar to that of the so called \emph{circular median} that is based on minimising arc length (without squaring)~\cite{Purkayastha_more_on_circ_median_1995,Liu_ordering_dir_data_1992,Chan_more_median_estimators_on_spheres_1993,mardia_stat_dir_data_jorver_1975,Mardia_directional_statistics,Otieno_phd_circ_2002,Otieno_circ_dir_ecolo_2006}. Surprisingly the \emph{unwrapped mean} has received little attention in the literature. An exception is Chan and He~\cite{Chan_more_median_estimators_on_spheres_1993} who briefly discuss the idea of minimising squared arc length, but focus more intently on minimising the arc length directly.  It appears that estimators based on minimising squared arc length have been overlooked. Perhaps one reason for this is that no efficient way of computing an estimate of the unwrapped mean from a sequence of observations was previously available.  We will describe an estimator called the \emph{angular least squares estimator} that can be computed efficiently using an algorithm from algebraic number theory for finding a \emph{nearest lattice point} in a famous lattice called $A_n^*$ \cite{McKilliam2009CoxeterLattices,McKilliam2008b,McKilliam2008,SPLAG,Conway1982FastQuantDec,Conway1982VoronoiRegions,Martinet2003,Voronoi1908_main_paper}. The number of operations required is $O(N)$, the same as required by the sample circular mean. We also describe the asymptotic properties of the angular least squares estimator showing that it is a strongly consistent estimator of the unwrapped mean and that it is asymptotically normally distributed.

Maximum likelihood estimation has been a common approach to direction estimation in the circular statistics literature (as it is in all of estimation theory)~\cite{Mardia_directional_statistics,Jammalamadaka_dir_stat_book}. Unfortunately, maximum likelihood estimation implicitly assumes that the underlying probability density function is known. For random variables on the real line a common assumption is that of Gaussianity and this is typically justified by appealing to the central limit theorem, but is sometimes also justified experimentally~\cite{Nyquist1924}. For circular random variables there is less evidence to suggest that a particular distribution should be chosen. A common assumption is that of the von Mises distribution~\cite{Upton_approx_cf_vm_1986} or the wrapped normal~\cite{Craig_time_ser_circ_thesis1988}, but such assumptions are not always convincing. This problem has been expressed by many authors and has encouraged the use of non-parametric techniques, such as bootstrapping, for dealing with circular statistics~\cite{RaoJammalamadaka_non_parimetric_bootstrap_circ_1984,Brunner_nonpari_circ_1994,Fisher_bootstrap_geo_1_1990}\cite[Ch.~7]{Jammalamadaka_dir_stat_book}\cite[Ch.~8]{Fisher1993}.

Unfortunately, these non-parametric techniques are typically computationally intensive and do not suit applications in signal processing such as phase and frequency estimation, polynomial phase estimation, noncoherent detection, or delay estimation that are often implemented on low-power computing devices and operate potentially thousands of times per second.  So our aim is to construct computationally efficient estimators that work well under a wide variety of distributional assumptions. In this paper we show that both the sample circular mean and the angular least squares estimator satisfy this goal.

The paper is organised in the following way.  In Section~\ref{sec:circ-rand-vari} we briefly introduce circular random variables and relevant notation. In Section~\ref{sec:circ-mean-vari} we describe the circular mean and show that the sample circular mean converges almost surely to the circular mean whenever it \emph{exists} (in a sense we will make precise). We show that the sample circular mean is asymptotically normally distributed. These results have appeared previously in the literature in various forms~\cite{Quinn2009_dasp_phase_only_information_loss,Fisher_common_mean_direction_dir_est_no_dist_assumptions1983,Jammalamadaka_dir_stat_book,mardia_stat_dir_data_book_1975}.  In Section~\ref{sec:unwr-mean-vari} we define the unwrapped mean and the angular least squares estimator. It is shown how the angular least squares estimator can be computed in linear-time by finding a nearest point in the lattice $A_n^*$. It is also shown that the angular least squares estimator converges almost surely to the unwrapped mean whenever it \emph{exists} and that it is asymptotically normally distributed. Similar results under stronger conditions were given by Quinn~\cite{Quinn2007}, but the proof we present is more general and takes a different approach.

The circular mean and the unwrapped mean are not always equal, but in Section~\ref{sec:relationships_circ_unwrapped_mean} we describe a large class of symmetric circular random variables for which they are equal.  For such circular random variables the angular least squares estimator and sample circular mean are estimators of the same quantity and can therefore be compared.  In Section~\ref{sec:comparingestimatorcircmean} we present simulations comparing the performance of the estimators under various distributional conditions.  The central limit theorems derived in Sections~\ref{sec:circ-mean-vari}~and~\ref{sec:unwr-mean-vari} lead to the hypothesis that the angular least squares estimator will be more accurate when the distribution is `light-tailed' whereas the sample circular mean estimator will be more accurate when the distribution is `heavy-tailed'. We have used the terms `light-tailed' and `heavy-tailed' loosely here, but it will become apparent how this is an appropriate and useful rule of thumb. The simulations are in agreement with the theoretical results and also support this `heavy-tail, light-tail' hypothesis.

@@

As the purpose of circular statistics is to describe the nature of \emph{angles} it is common in the literature to define circular random variables to take values on $[0, 2\pi)$ or $[-\pi, \pi)$. In this paper we find it more convenient to define circular random variables to take values on the interval $[-\nicefrac{1}{2}, \nicefrac{1}{2})$. So, when we refer to an \emph{angle} we mean a real number in the interval $[-\nicefrac{1}{2}, \nicefrac{1}{2})$. This is nonstandard but it will allow us to use notation such as $\round{\cdot}$ for rounding and $\fracpart{\cdot}$ for the fractional part in a convenient way, and will also lead to close ties between circular statistics and number theory or, more specifically, the lattice $A_n^*$.

It is common in the literature to define a special \emph{circular} probability density function (pdf) $f$ to be periodic with period $1$ (or $2\pi$) so that $f(\theta + k) = f(\theta)$ for any integer $k$ and the integral $\int_{T}f(\theta)d\theta = 1$ where $T$ is any interval of length one.  We will not use this definition.  In this paper a circular random variable is a random variable with range within $[-\nicefrac{1}{2},\nicefrac{1}{2})$. The utility of this is that we have not separated \emph{circular} random variables from \emph{regular} random variables in any way. Sometimes it will be convenient to think of a circular random variable as a random variable that returns values in $[-\nicefrac{1}{2},\nicefrac{1}{2})$ and other times it will be more natural to think of a circular random variable as describing angles wrapped around a circle.  %By not making any severe distinction between circular random variables and regular random variables we are able to switch between these two concepts freely without any notational baggage.  
If we wish to consider the pdf $f$ as a periodic function we shall use $f(\fracpart{x})$, which is clearly periodic with period one for $x \in \reals$ because the fractional part function $\fracpart{x} = x - \round{x}$ has period one.

A circular random variable inherits all the properties of a \emph{regular} random variable.  For example, if $\Theta$ is a circular random variable with pdf $f$ the expected value of a function $g(\Theta)$ of $\Theta$ is given in the usual way by
\[
E[g(\Theta)] = \int_{-\infty}^{\infty}g(\theta)f(\theta)d\theta = \int_{-\nicefrac{1}{2}}^{\nicefrac{1}{2}}g(\theta)f(\theta)d\theta.
\]
This leads to the usual definitions of mean and variance for a circular random variable $E[\Theta]$ and $\var[\Theta] = E[\Theta^2] - E[\Theta]^2$.  A little thought must be given here. The mean $E[\Theta]$ does not necessarily correspond to the \term{mean direction} of $\Theta$ in the sense one might expect.  For example, consider a distribution with equal point masses at $-0.49$ and $0.49$.  The expected value of this random variable is zero, but intuitively a more reasonable estimate for the mean direction is $-\nicefrac{1}{2}$.  One property that any definition of mean direction should have is invariance to translation modulo 1.  That is, if $\Theta$ is a circular random variable with mean direction $\mu$ then for any $\phi \in \reals$ the translated (modulo 1) circular random variable $\fracpart{\Theta + \phi}$ should have mean direction $\fracpart{\mu + \phi}$.  In the following sections we consider two different quantities called the \term{circular mean} and the \term{unwrapped mean} that satisfy this invariance property and correspond with our intuitive notion of mean direction.

%In this paper we will quite often use the mean and variance of a circular random variable i.e. $E[\Theta]$ and $\var[\Theta]$.  When we want one of the other notions, i.e. the circular mean and circular variance, or the unwrapped mean and unwrapped variance we will always state these names in full.  This is in contrast to much of the circular statistics literature that uses the terms mean and variance to refer to the circular mean and circular variance.

@@

Given a circular random variable with pdf $f$ the most common analogue of `mean' and `variance' in the literature is the \term{circular mean} given by 
\[
\mu_{\text{circ}} = \frac{\angle{c}}{2\pi}
\] 
and the \term{circular variance} given by 
\[
\nu = 1 - |c|
\]  
where 
\begin{equation}\label{eq:circmeanint}
c = \int_{-\nicefrac{1}{2}}^{\nicefrac{1}{2}} e^{2\pi j \theta} f(\theta) d\theta,
\end{equation} 
and $\angle{c}$ and $|c|$ are respectively the complex argument and the magnitude of $c$ \cite[p.~29]{Mardia_directional_statistics}\cite{Fisher1993}. Consider the case when $c = 0$. The circular variance is $1$ but the circular mean is undefined.  In this paper we will say that the distribution \emph{has no circular mean} when $c = 0$. %This is contrast to the mean of a random variable, which is given by $E[\Theta]$ and is always defined.

@@

%\begin{figure}[tbp]
%	\centering
%		\includegraphics[height=6cm]{apps_anstar/distplots/initialdistpic.1}
%		\caption{A bimodal circular probability density function. The circular and unwrapped means are marked. They are not equal.}
%		\label{fig:plot_initcirc_bimodal}
%\end{figure}

\begin{figure}[tp]
	\centering 
		\includegraphics[width=\linewidth]{figs/wrappeduniform_var0p08-1.mps}
		\caption{The circular uniform distribution which has no circular mean and no unwrapped mean.}
		\label{fig:circularuniformdist}
\end{figure}

\begin{figure}[tp]
	\centering
		\includegraphics[width=\linewidth]{figs/circnomeansym-1.mps}
		\caption{A bimodal distribution with no circular mean and no unwrapped mean.}
		\label{fig:circnomeanbimodal}
\end{figure}

@@

\section{The unwrapped mean and its estimation}\label{sec:unwr-mean-vari}

>>>>

\author{Robby~G.~McKilliam, Barry G. Quinn and I. Vaughan L. Clarkson%
  \thanks{A preliminary version of some of this material is contained in Chapters 5 and 6 of Robby McKilliam's PhD thesis \cite{McKilliam2010thesis}. Robby~McKilliam is with the Institute for Telecommunications Research, The University of South Australia, SA, 5095.  Barry~Quinn is with the Department of Statistics, Macquarie University, Sydney, NSW, 2109, Australia.   Vaughan~Clarkson is with the School of Information Technology \& Electrical Engineering, The University of Queensland, QLD., 4072, Australia.}
     }
% The paper headers 
\markboth{Direction estimation by minimum squared arc length}{DRAFT \today}

@@

\begin{abstract}
Circular statistics has found substantial application in science and engineering. One of the fundamental problems in circular statistics is that of estimating the \emph{mean direction} of a circular random variable from a number of observations. %We consider to different definitions of the mean direction, the \emph{circular} (or \emph{extrinsic}) \emph{mean} and the \emph{intrinsic mean}. 
The standard approach in the literature is called the \emph{sample circular mean} and its asymptotic properties are well known. It can also be computed efficiently in a number of arithmetic operations that is linear in the number of observations. In this paper we consider an alternative estimator called the \emph{sample intrinsic mean} that is based on minimising squared arc length. We show how this estimator can be computed efficiently in a linear number of operations using an algorithm from algebraic number theory and we derive its asymptotic properties. 
In some scenarios the sample circular mean and the sample intrinsic mean are estimators of the same quantity and can therefore be compared.  We show both theoretically and by simulation that in some of these scenarios the sample intrinsic mean is statistically more accurate than the sample circular mean. As such the results in this paper potentially have implications for the wide variety of fields in science, engineering and statistics that currently use the sample circular mean.
\end{abstract}

\begin{IEEEkeywords}
Circular statistics, mean direction estimation, intrinsic mean, lattice theory, nearest lattice point problem.
\end{IEEEkeywords}

@@

The field of circular statistics aims to describe the nature of data that is measured in angles or 2-dimensional unit vectors or complex numbers on the unit circle \cite{Mardia_directional_statistics,Jupp_mardia_unified_directional_statistics_1989,mardia_stat_dir_data_book_1975,Fisher1993,Jammalamadaka_dir_stat_book}.  Such data occur frequently in science, particularly in astronomy, biology \cite{batschelet1981circular,Cochranmigbridcalcircstat2004,Bolesspinylobnavcircstat2003}, medicine \cite{Mann_circstat_med_cement_2003,Lediseasecircstat2003}, geology, geography and meteorology~\cite{Fisher_bootstrap_geo_1_1990,Marida_geosciences_1981}, and also in engineering, particularly in communications and radar \cite{Fogel1989_bit_synch_zero_crossings,Elton96robustparameter1996,Clarkson2007,McKilliamLinearTimeBlockPSK2009,McKilliam2007}.  A meteorological example is the direction of the wind, and a biological example is the direction of flight taken by a bird. %The field of circular statistics is in some sense an ancient one, probably begun when mankind first started recording the motion of the sun, the moon and the stars.  A thorough historical account of the subject is given by Fisher \cite[Ch.~1]{Fisher1993}.
Circular statistics can be considered as a special case of the more general field of statistics on Riemannian manifolds~\cite{Bhattacharya_int_ext_means_2003,Bhattacharya_int_ext_means_2005,Pennec_int_st_rm_2006,bwhk07a}.  Circular statistics is then the study of statistics on the manifold $S^1$.  Here we focus specifically on circular statistics but we will make some connections to the more general theory.

A fundamental and useful problem in circular statistics is that of estimating the \emph{mean direction} of a circular random variable from a number, say $N$, of observations. For example, if you listen to the weather report you may be told (an estimate of) the direction of the wind.  Obtaining an accurate estimate requires a method for accurately estimating the mean wind direction from a number of observations of the wind direction. The problem of estimating mean direction also appears in signal processing problems such as phase and frequency estimation \cite{Quinn2009_dasp_phase_only_information_loss,Lovell1991,McKilliamFrequencyEstimationByPhaseUnwrapping2009,Quinn2001,McKilliam2010thesis}, polynomial phase estimation~\cite{McKilliam2009asilomar_polyest_lattice,McKilliam2010thesis,Kitchen_polyphase_unwrapping_1994}, noncoherent detection \cite{McKilliamLinearTimeBlockPSK2009,McKilliam2010thesis,Mackenthun1994} and delay and period estimation from sparse observations \cite{Elton96robustparameter1996,Clarkson2007,McKilliam2007,McKilliam2010thesis,Fogel1989_bit_synch_zero_crossings}.

Unlike the case of random variables on the real line where the mean is uniquely defined, the definition of \emph{mean direction} of a circular random variable is somewhat open to interpretation. The most common definition in the literature, called the \term{circular mean}, is derived by considering circular random variables as unit vectors in 2-dimensions or equivalently complex numbers on the unit circle. Given $N$ observations the appropriate estimator for the circular mean is the \emph{sample circular mean} which can be computed efficiently by averaging $N$ complex numbers. This requires only $O(N)$ arithmetic operations.  The asymptotic properties of the sample circular mean under various assumptions are also well studied \cite{Mardia_directional_statistics,Quinn2009_dasp_phase_only_information_loss,Jammalamadaka_dir_stat_book,McKilliam2010thesis}.  In the more general theory of statistics on manifolds the circular mean is called the \emph{extrinsic mean} and its asymptotic properties are know even in this more general setting~\cite{Bhattacharya_int_ext_means_2003}.

In this paper we consider an alternative definition of mean direction called the \term{intrinsic mean}\index{intrinsic mean} which is based on minimising the expected squared arc length. %This definition is similar to that of the so called \emph{circular median} that is based on minimising arc length (without squaring)~\cite{Purkayastha_more_on_circ_median_1995,Liu_ordering_dir_data_1992,Chan_more_median_estimators_on_spheres_1993,mardia_stat_dir_data_jorver_1975,Mardia_directional_statistics,Otieno_phd_circ_2002,Otieno_circ_dir_ecolo_2006}.  
%In the theory of statistics on manifolds 
The intrinsic mean is sometimes also called the \emph{Fr\'{e}chet mean}.  The intrinsic mean can be estimated by the \emph{sample intrinsic mean} and it is known that this estimator is strongly consistent~\cite{Ziezold_intrinsic_mean_1977,Bhattacharya_int_ext_means_2003}.  The asymptotic distribution of the intrinsic mean has been studied by Bhattacharya and  Patrangenaru~\cite{Bhattacharya_int_ext_means_2005} under restrictive assumptions on the underlying probability distribution.  For the purpose of circular statistics these assumptions involve the distribution having no mass in an interval containing the point antipodal (or opposite) to the intrinsic mean.  This precludes many popular circular distributions such as the von Mises, the wrapped normal, and the projected normal.

Restricting to the case of circular statistics Quinn~\cite{Quinn2007} relaxed these assumption and derived the asymptotic distribution of the sample intrinsic mean for the case when the probability distribution is unimodal and symmetric.  Here, we relax these unimodality assumptions.  We derive the asymptotic distribution of the sample intrinsic mean showing that it is asymptotically normally distributed whenever the underlying distribution has a well defined intrinsic mean.
% and satisfies a mild condition on the value of its density function at the point antipodal to the intrinsic mean. BLERG

%Surprisingly the \emph{intrinsic mean} has received little attention in the literature. An exception is Chan and He~\cite{Chan_more_median_estimators_on_spheres_1993} who briefly discuss the idea of minimising squared arc length, but focus more intently on minimising the arc length directly.  It appears that estimators based on minimising squared arc length have been overlooked. Perhaps one reason for this is that no efficient way of computing an estimate of the intrinsic mean from a sequence of observations was previously available.

Computing the sample intrinsic mean is generally considered to be a difficult problem. Bhattacharya and Patrangenaru~\cite{Bhattacharya_int_ext_means_2005} state, ``Much of the literature deals with the special case of what we call the extrinsic mean, perhaps because of the technical difficulties involved in proving the existence of an intrinsic mean and in computing the intrinsic sample mean even when it exists''.  Here we will show that for the specific case of circular statistics the sample intrinsic mean can be computed efficiently using an algorithm from algebraic number theory for finding a \emph{nearest lattice point} in a lattice called $A_n^*$ \cite{McKilliam2009CoxeterLattices,McKilliam2008b,McKilliam2008,SPLAG,Conway1982FastQuantDec,Martinet2003}. The number of operations required is $O(N)$, the same as required by the sample circular mean.

%Maximum likelihood estimation has been a common approach to direction estimation in the circular statistics literature (as it is in all of estimation theory)~\cite{Mardia_directional_statistics,Jammalamadaka_dir_stat_book}. Unfortunately, maximum likelihood estimation implicitly assumes that the underlying probability density function is known. For random variables on the real line a common assumption is that of Gaussianity and this is typically justified by appealing to the central limit theorem, but is sometimes also justified experimentally~\cite{Nyquist1924}. For circular random variables there is less evidence to suggest that a particular distribution should be chosen. A common assumption is that of the von Mises distribution~\cite{Upton_approx_cf_vm_1986} or the wrapped normal~\cite{Craig_time_ser_circ_thesis1988}, but such assumptions are not always convincing. This problem has been expressed by many authors and has encouraged the use of non-parametric techniques, such as bootstrapping, for dealing with circular statistics~\cite{RaoJammalamadaka_non_parimetric_bootstrap_circ_1984,Brunner_nonpari_circ_1994,Fisher_bootstrap_geo_1_1990}\cite[Ch.~7]{Jammalamadaka_dir_stat_book}\cite[Ch.~8]{Fisher1993}.

%Unfortunately, these non-parametric techniques are typically computationally intensive and do not suit applications in signal processing such as phase and frequency estimation, polynomial phase estimation, noncoherent detection, or delay estimation that are often implemented on low-power computing devices and operate potentially thousands of times per second.  So our aim is to construct computationally efficient estimators that work well under a wide variety of distributional assumptions. In this paper we show that both the sample circular mean and the sample intrinsic mean satisfy this goal.

The paper is organised in the following way.  In Section~\ref{sec:circ-rand-vari} we briefly introduce circular random variables and relevant notation. In Section~\ref{sec:circ-mean-vari} we describe the circular mean and show that the sample circular mean converges almost surely to the circular mean whenever it \emph{exists} (in a sense we will make precise). We show that the sample circular mean is asymptotically normally distributed. These results have appeared previously in the literature in various forms~\cite{Quinn2009_dasp_phase_only_information_loss,Fisher_common_mean_direction_dir_est_no_dist_assumptions1983,Jammalamadaka_dir_stat_book,mardia_stat_dir_data_book_1975,Bhattacharya_int_ext_means_2003}.  In Section~\ref{sec:unwr-mean-vari} we define the intrinsic mean and the sample intrinsic mean. It is shown how the sample intrinsic mean can be computed in linear-time by finding a nearest point in the lattice $A_n^*$. It is also shown that the sample intrinsic mean converges almost surely to the intrinsic mean whenever it exists and that it is asymptotically normally distributed. %Similar results under stronger conditions were given by Quinn~\cite{Quinn2007}, but the proof we present is more general and takes a different approach.

The circular mean and the intrinsic mean are not always equal, but in Section~\ref{sec:relationships_circ_intrinsic_mean} we describe a large class of symmetric circular random variables for which they are equal.  For such circular random variables the sample intrinsic mean and sample circular mean are estimators of the same quantity and can be meaningfully compared.  In Section~\ref{sec:comparingestimatorcircmean} we present simulations comparing the performance of the estimators under various distributional conditions.  The central limit theorems derived in Sections~\ref{sec:circ-mean-vari}~and~\ref{sec:unwr-mean-vari} lead to the hypothesis that the sample intrinsic mean will be more accurate when the distribution is `light-tailed' whereas the sample circular mean will be more accurate when the distribution is `heavy-tailed'. We have used the terms `light-tailed' and `heavy-tailed' loosely here, but it will become apparent how this is an appropriate and useful rule of thumb. The simulations are in agreement with the theoretical results and also support this `heavy-tail, light-tail' hypothesis.

@@

As the purpose of circular statistics is to describe the nature of angles it is common in the literature to define circular random variables to take values on $[0, 2\pi)$ or $[-\pi, \pi)$. In this paper we find it more convenient to define circular random variables to take values on the interval $[-\nicefrac{1}{2}, \nicefrac{1}{2})$. So, when we refer to an angle we mean a real number in the interval $[-\nicefrac{1}{2}, \nicefrac{1}{2})$. This is nonstandard but it will allow us to use notation such as $\round{\cdot}$ for rounding and $\fracpart{\cdot}$ for the fractional part in a convenient way, and will also lead to close ties between circular statistics and number theory or, more specifically, the lattice $A_n^*$.

It is common in the literature to define a special \emph{circular} probability density function (pdf) $f$ to be periodic with period $1$ (or $2\pi$) so that $f(\theta + k) = f(\theta)$ for any integer $k$ and the integral $\int_{T}f(\theta)d\theta = 1$ where $T$ is any interval of length one.  We will not use this definition.  In this paper a circular random variable is a random variable that takes values in $[-\nicefrac{1}{2},\nicefrac{1}{2})$. The utility of this is that we have not separated \emph{circular} random variables from \emph{regular} random variables in any way. Sometimes it will be convenient to think of a circular random variable as a random variable that returns values in $[-\nicefrac{1}{2},\nicefrac{1}{2})$ and other times it will be more natural to think of a circular random variable as describing angles wrapped around a circle.  %By not making any severe distinction between circular random variables and regular random variables we are able to switch between these two concepts freely without any notational baggage.  
If we wish to consider the pdf $f$ as a periodic function we shall use $f(\fracpart{x})$, which is clearly periodic with period one for $x \in \reals$ because the fractional part function $\fracpart{x} = x - \round{x}$ has period one.

A circular random variable inherits all the properties of a regular random variable.  For example, if $\Theta$ is a circular random variable with pdf $f$ the expected value of a function $g(\Theta)$ of $\Theta$ is given in the usual way by
\[
E[g(\Theta)] = \int_{-\infty}^{\infty}g(\theta)f(\theta)d\theta = \int_{-\nicefrac{1}{2}}^{\nicefrac{1}{2}}g(\theta)f(\theta)d\theta.
\]
This leads to the usual definitions of mean and variance for a circular random variable $E[\Theta]$ and $\var[\Theta] = E[\Theta^2] - E[\Theta]^2$.  A little thought must be given here. The mean $E[\Theta]$ does not necessarily correspond to the \term{mean direction} of $\Theta$ in the sense one might expect.  For example, consider a distribution with equal point masses at $-0.49$ and $0.49$.  The expected value of this random variable is zero, but intuitively a more reasonable estimate for the mean direction is $-\nicefrac{1}{2}$.  One property that any definition of mean direction should have is invariance to translation modulo 1.  That is, if $\Theta$ is a circular random variable with mean direction $\mu$ then for any $\phi \in \reals$ the translated (modulo 1) circular random variable $\fracpart{\Theta + \phi}$ should have mean direction $\fracpart{\mu + \phi}$.  In the following sections we consider two different quantities called the \term{circular mean} and the \term{intrinsic mean} that satisfy this invariance property and correspond with our intuitive notion of mean direction.

%In this paper we will quite often use the mean and variance of a circular random variable i.e. $E[\Theta]$ and $\var[\Theta]$.  When we want one of the other notions, i.e. the circular mean and circular variance, or the intrinsic mean and unwrapped variance we will always state these names in full.  This is in contrast to much of the circular statistics literature that uses the terms mean and variance to refer to the circular mean and circular variance.

@@

Given a circular random variable with pdf $f$ the most common analogue of `mean' and `variance' in the literature is the \term{circular mean} given by 
\[
\mu_{\text{circ}} = \frac{\angle{c}}{2\pi}
\] 
and the \term{circular variance} given by 
\[
\nu = 1 - |c|,
\]  
where 
\begin{equation}\label{eq:circmeanint}
c = \int_{-\nicefrac{1}{2}}^{\nicefrac{1}{2}} e^{2\pi j \theta} f(\theta) d\theta,
\end{equation} 
and $j = \sqrt{-1}$, and $\angle{c}$ and $|c|$ are respectively the complex argument and the magnitude of $c$ \cite[p.~29]{Mardia_directional_statistics}\cite{Fisher1993}. Consider the case when $c = 0$. The circular variance is $1$ but the circular mean is undefined.  We say that the distribution \emph{has no circular mean} when $c = 0$. %This is contrast to the mean of a random variable, which is given by $E[\Theta]$ and is always defined.

@@

%\begin{figure}[tbp]
%	\centering
%		\includegraphics[height=6cm]{apps_anstar/distplots/initialdistpic.1}
%		\caption{A bimodal circular probability density function. The circular and intrinsic means are marked. They are not equal.}
%		\label{fig:plot_initcirc_bimodal}
%\end{figure}

\begin{figure}[tp]
	\centering 
		\includegraphics[width=\linewidth]{figs/wrappeduniform_var0p08-1.mps}
		\caption{The circular uniform distribution which has no circular mean and no intrinsic mean.}
		\label{fig:circularuniformdist}
\end{figure}

\begin{figure}[tp]
	\centering
		\includegraphics[width=\linewidth]{figs/circnomeansym-1.mps}
		\caption{A bimodal distribution with no circular mean and no intrinsic mean.}
		\label{fig:circnomeanbimodal}
\end{figure}

@@

\section{The intrinsic mean and its estimation}\label{sec:unwr-mean-vari}

Alternatives to the circular mean and circular variance are the \term{intrinsic mean} and \term{intrinsic variance}.  
%We have chosen the term `unwrapped' because the original motivation for the intrinsic mean and the sample intrinsic mean came from the \emph{phase unwrapping} literature, in particular the problems of \emph{frequency estimation} and \emph{polynomial phase estimation} \cite{Tretter1985,McKilliamFrequencyEstimationByPhaseUnwrapping2009,Quinn_on_kay_2000,Morelande_bayes_unwrapping_2009_tsp,Fowler2002_freq_est_by_phases,McKilliam_polyphase_est_icassp_2011}. The idea behind the sample intrinsic mean is to `unwrap' the circle as will be seen in the next section. Let $\Theta$ be a circular random variable.  
The intrinsic mean of $\Theta$ is the angle $\mu_{\text{intr}}$ such that the translated circular random variable $\fracpart{\Theta - \mu_{\text{intr}}}$ has minimum squared expectation, that is,
\begin{align}
\mu_{\text{intr}}  &= \arg \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})}E\fracpart{\Theta - \mu}^2 \nonumber \\
&= \arg \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})}\int_{-\nicefrac{1}{2}}^{\nicefrac{1}{2}}\fracpart{\theta - \mu}^2f(\theta) d\theta. \label{eq:minunrappedmeandef}
\end{align}
The intrinsic variance of $\Theta$, denoted $\sigma^2$, is given by 
\[
\sigma^2 = E\fracpart{\Theta - \mu_{\text{intr}}}^2 = \int_{-\nicefrac{1}{2}}^{\nicefrac{1}{2}}\fracpart{\theta - \mu_{\text{intr}}}^2f(\theta) d\theta.
\]
%Intuitively the intrinsic mean is such that the pdf of $\fracpart{\Theta - \mu_{\text{intr}}}$ is mostly centred around zero.

For some circular distributions the minimisation that defines the intrinsic mean~(\ref{eq:minunrappedmeandef}) might not be unique\footnote{The set of minima is often called the~\emph{Fr\'{e}chet mean set}~\cite{Bhattacharya_int_ext_means_2003,Bhattacharya_int_ext_means_2005}.}.  For example, consider the bimodal probability density function depicted in Figure~\ref{fig:circnomeanbimodal} for which there would by two minima, one at $\nicefrac{1}{2}$ and one at $-\nicefrac{1}{2}$ (it may at first seem that the minima would be at the modes $\nicefrac{1}{4}$ and $-\nicefrac{1}{4}$, but this is not true in this case).  Another example is the circular uniform distribution (Figure~\ref{fig:circularuniformdist}) where the integral from~\eqref{eq:minunrappedmeandef} is $\nicefrac{1}{12}$ for all $\mu$.  We say that such distributions have no intrinsic mean.

% A circular random variable, $\Theta$, with zero intrinsic mean, i.e. $\mu_{\text{intr}} = 0$, has the special property that the intrinsic mean is equal to the mean, that is
% \begin{equation}\label{eq:intrmean0=mean}
% \mu_{\text{unwrap}} = E[\Theta] = 0
% \end{equation}
% and the unwrapped variance is equal to the variance, that is
% \begin{equation}\label{eq:unwrapvar0=var}
% \sigma^2 = \var[\Theta].
% \end{equation}
